### 2024-07-01 causal-inference.qmd
20240630 - genetically informed designs

>>>>

Another approach to control for variables is to use genetically informed designs.
Genetically informed designs allow controlling for potential genetic effects in order to more closely approximate the contributions of various environmental effects.
Genetically informed designs exploit differing degrees of genetic relatedness among participants to capture the extent to which genetic factors may contribute to an outcome.
The average percent of DNA shared between people of varying relationships is provided in @tbl-geneticRelatednessRelativePairs (<https://isogg.org/wiki/Autosomal_DNA_statistics>; archived at <https://perma.cc/MK3D-DST8>):

| Relationship                       | Average Percent of Autosomal DNA Shared by Pairs of Relatives |
|:-----------------------------------|:--------------------------------------------------------------|
| Monozygotic ("identical") twins    | 100%                                                          |
| Dizygotic ("fraternal") twins      | 50%                                                           |
| Parent/child                       | 50%                                                           |
| Full siblings                      | 50%                                                           |
| Grandparent/grandchild             | 25%                                                           |
| Aunt-or-uncle/niece-or-nephew      | 25%                                                           |
| Half-siblings                      | 25%                                                           |
| First cousin                       | 12.5%                                                         |
| Great-grandparent/great-grandchild | 12.5%                                                         |

: Average Percent of Autosomal DNA Shared by Pairs of Relatives by Relationship Type. {#tbl-geneticRelatednessRelativePairs}

For instance, researchers may compare monozygotic twins versus dizygotic twins in some outcome—a so-called "twin study".
It is assumed that the trait/outcome is attributable to genetic factors to the extent that the monozygotic twins (who share 100% of their DNA) are more similar in the trait or outcome compared to the dizygotic twins (who share on average 50% of their DNA).
Alternatively, researchers could compare full siblings versus half-siblings, or they could compare full siblings versus first cousins.

Genetically informed designs are not as relevant for fantasy football analytics, but they are useful to present as one of various design features that researchers can draw upon to strengthen their ability to make causal inferences.
